Practical Optional Types for Clojure

### 2015-07-09 overview.tex
rewrite middle of section 3
Add feeds2imap experiment
Edits to 2.

Here is a simple program in Typed Clojure.

We define a global variable \clj{greet} as a one-argument function
taking and returning a string.

@@

\begin{minted}{clj}
(greet 
  nil ; Type Error:
      ;   Expected Str, given nil
  )
\end{minted}

@@

\paragraph{Flow analysis} Typed Clojure uses occurrence typing to
model type-based control flow.
In \clj{greetings}, a \clj{when} expression ensures \clj{repeat}
is never passed a \clj{nil} argument.

@@

\begin{minted}{clj}
  (.getParent (new File "a/b"))  ; :- (U nil Str)
  ;=> "a"
\end{minted}

@@

\begin{minted}{clj}
  (fn [f] 
    (.getParent f) ; Type Error:
                   ;   Unresolved interop: getParent
    )
\end{minted}

Function arguments default to \clj{Any}, the most permissive type. Ascribing
a parameter type helps Typed Clojure find a specific method.

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.

\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f) ; Type Error:
                 ;   Cannot call instance method 
                 ;   java.io.File/getParent on type 
                 ;   (U nil File).
  )
\end{minted}

@@

\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f) ; Type Error:
                 ;   Expected Str, given (U nil Str).
  )
\end{minted}

@@

By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
but can be configured globally for particular non-target positions if needed.

@@

A \emph{multimethod} is a kind of extensible function. It combines
a \emph{dispatch function} with one or more \emph{methods}.

\paragraph{Value-based dispatch}
Here is a simple multimethod.

\begin{minted}{clojure}
(ann hi [Kw -> Str]) ; interface type for `hi`
(defmulti hi identity) ; dispatch function `identity`
(defmethod hi :en [_] "hello") ; method for `:en`
(defmethod hi :fr [_] "bonjour") ; method for `:fr`
(defmethod hi :default [_] "um...") ; default method
\end{minted}

@@

This is only part of the story for multimethod dispatch.

\paragraph{Class-based dispatch}
For class values, multimethods choose methods based on subclassing
relationships.

Recall the multimethod in figure~\ref{fig:ex1}, duplicated below.

\begin{minted}{clj}
(ann pname [(U File String) -> (U nil String)])
(defmulti pname class)
(defmethod pname String [s] (pname (new File s)))
(defmethod pname File [f] (.getName f))
\end{minted}

%Its dispatching function is
%\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
%respectively.

The dispatch function \clj{class}
%---associated at multimethod creation with \clj{defmulti}---
dictates 
whether the \clj{String} or \clj{File} method is chosen.
%---both installed via \clj{defmethod}

The multimethod dispatch rules involve
\clj{isa?}, a hybrid predicate which includes a subclassing check for classes and
an equality check for most other values.

@@

To demonstrate,
  \clj{(pname "STAINS/JELLY")}
chooses the \clj{String} method because
\clj{(isa? (class "STAINS/JELLY") String)}
is true and
\clj{(isa? (class "STAINS/JELLY") File)}
is false. Then the \clj{String} method body
\clj{(pname (new File "STAINS/JELLY"))}
chooses the \clj{File} method for opposite reasons,
resulting in 
\begin{minted}{clojure}
(.getName (new File "STAINS/JELLY")) ; :- (U nil Str)
;=> "JELLY"
\end{minted}

@@

Immutable hash-maps with keyword keys play a major role in Clojure programming
as the primary way to structure data.

\begin{minted}{clojure}
(def breakfast {:en "waffles" :fr "croissants"})

(get breakfast :en)    ; :- Str
;=> "waffles"
; keywords are functions that look themselves up
(:fr breakfast)        ; :- Str
;=> "croissants"
(:bocce breakfast) ; :- nil
;=> nil
\end{minted}

HMap types model the most common usages of keyword maps.
\begin{minted}{clojure}
breakfast ; :- (HMap :mandatory {:en Str :fr Str}
          ;          :complete? true)
\end{minted}
The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
and \clj{:fr} have string values (expressed as a \clj{:mandatory} entry) and that 
every other key is absent (by \clj{:complete?} being \clj{true}).

HMap's default to being partially specified---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
\clj{'{:en Str :fr Str}}
forgets 
information about absent keys, only providing \clj{:mandatory}
information.

@@

\paragraph{HMaps in practice} The next example is extracted from a production system at CircleCI,
a company with a large production Typed Clojure system
(section~\ref{sec:casestudy} presents a case study).

\begin{minted}{clojure}
(defalias RawKeyPair
  "Unencrypted keypair -- extra keys disallowed"
  (HMap :mandatory {:public-key RawKey,
                    :private-key RawKey},
        :complete? true))

@@

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair "Encrypt an unencrypted keypair"
  [{pk :private-key :as kp}] ; original map is kp
  (assoc 
    ; remove unencrypted private key
    (dissoc kp :private-key)
    ; add encrypted private key
    :enc-private-key (encrypt pk)))
\end{minted}
%\begin{exmp}
%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
%\label{example:circleci}
%\end{exmp}
Fully specified HMap's
statically enforce the private key is not accidentally left in a supposedly
encrypted keypair.

\begin{minted}{clojure}
(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
(defn enc-keypair-bad
  [{pk :private-key :as kp}]
  (assoc kp :enc-private-key (encrypt pk))
  ; Type Error:
  ;   Expected EncKeyPair, given 
  ;   (HMap :mandatory {:enc-private-key EncKey
  ;                     :private-key RawKey
  ;                     :public-key RawKey}
  ;         :complete? true)
  )
\end{minted}

The extra \clj{:private-key} entry disagrees with \clj{EncKeyPair}, so a type error
is given.

@@

Since HMaps are the primary way of specifying the structure of data in Clojure,
and multimethods are the primary tool for dispatching on data, they are inevitably
linked.
There are infinite ways of both structuring and dispatching on data, so there is no
hope predicting them all, or even manually covering useful a subset.

Thankfully, occurrence typing is both extensible \emph{and} compositional.
By extending occurrence typing with
a handful of rules based on HMaps and other functions, 
we can automatically cover almost all the common cases---as well others
that compose simple rules in arbitrary ways.

Futhermore, no special attention is needed for multimethod dispatch---the primitive branching
mechanism is still the humble \clj{if} conditional. Only a small number of rules are needed
to encode the \clj{isa?}-based dispatch, themselves made of small, simple pieces.
In practice, this means all conditional-based control flow analysis
also extends to the context of multimethod dispatch, and vice-versa.

@@

The \clj{:combo} method is verified to only structurally recur
on \clj{Order}s. This is achieved because we learn the argument--- \clj{o} ---must % em dash needs space here, looks weird
be of type
\clj{'{:Meal :combo}}
since
\clj{(isa? (:Meal o) :combo)}
must be true. Combining \clj{'{:Meal :combo}}
with the fact that \clj{o} is an \clj{Order}
eliminates possibility of \clj{:lunch} and \clj{:dinner}
orders, simplifying \clj{o} to
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
which can safely structurally recur.

@@

In the \clj{Long} method, Typed Clojure learns that
its argument is at least of type \clj{'{:desserts Long}}---since
\begin{minted}{clojure}
(isa? (class (:desserts o)) Long)
\end{minted}
must be true.

Knowing \clj{o} is also an
\clj{Order}, we can eliminate the possibility of
a \clj{:combo} meal, and infer the type
\begin{minted}{clojure}
(U '{:Meal ':lunch :desserts Int}
   '{:Meal ':dinner :desserts Int})
\end{minted}
Looking up \clj{:desserts} on this type returns an \clj{Int}.

In the \clj{nil} method, we know
\begin{minted}{clojure}
(isa? (class (:desserts o)) nil)
\end{minted}
is true---which implies \clj{(class (:desserts o))} is \clj{nil}.

Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
  \item \clj{o} is missing a \clj{:desserts} entry.
\end{itemize}
Equivalently, we learn \clj{o} is at least of type
\begin{minted}{clojure}
(U '{:desserts nil}
   ; :absent-keys, a set of known absent entries
   (HMap :absent-keys #{:desserts}))
\end{minted}
This intersected with \clj{o} being an \clj{Order} simplifies to a \clj{:combo} meal, 
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
thus verifying both structural recursions.

\paragraph{From multiple to arbitrary dispatch}
Multimethod dispatch goes even further, supporting
multiple dispatch via vectors.

Fully specifying multimethod dispatch is beyond the scope of this paper,
but the same intuition applies---adding support for multiple dispatch
automatically allows arbitrary combinations and nestings
of it and previous simple dispatch rules.

>>>>

Here is a simple program in Typed Clojure.
%
We define \clj{greet} as a one-argument function
taking and returning a string.

@@

\begin{minted}{clj}
(greet nil) ; Type Error:
            ;   Expected Str, given nil
\end{minted}

@@

\paragraph{Flow analysis} Typed Clojure uses occurrence typing~\cite{thf10} to
model type-based control flow.
In \clj{greetings}, a \clj{when} expression ensures \clj{repeat}
is never passed a \clj{nil} argument.

@@

\begin{exmp}
\begin{minted}{clj}
  (.getParent (new File "a/b"))  ; :- (U nil Str)
  ;=> "a"
\end{minted}
\label{example:getparent-direct-constructor}
\end{exmp}

@@

\begin{minted}{clj}
  (fn [f]           ; Type Error:                    
    (.getParent f)) ;   Unresolved interop: getParent

\end{minted}

Function arguments default to \clj{Any}, the most permissive type. Ascribing
a parameter type allows Typed Clojure to find a specific method.

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)) ; Type Error:
                  ;   Cannot call instance method 
                  ;   java.io.File/getParent on type 
                  ;   (U nil File).
\end{minted}
Since Java-level types do not provide information about nullability,
by default Typed Clojure conservatively assumes that Java method and
constructor arguments, such as to @clj{.getParent}
 must be \emph{non-nullable}, but this can be
configured for  particular calls as needed.

@@

\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f)) ; Type Error:
                  ;   Expected Str, given (U nil Str).
\end{minted}

@@

@@

\emph{Multimethods} are a kind of extensible function, and they are
widely used to define Clojure operations.  It combines a
\emph{dispatch function} with one or more \emph{methods}.

\paragraph{Value-based dispatch}
This simple multimethod says hello in different languages, as
specified by a keyword argument.

\begin{exmp}
\begin{minted}{clojure}
(ann hi [Kw -> Str]) ; interface type for `hi`
(defmulti hi identity) ; dispatch function `identity`
(defmethod hi :en [_] "hello") ; method for `:en`
(defmethod hi :fr [_] "bonjour") ; method for `:fr`
(defmethod hi :default [_] "um...") ; default method
\end{minted}
\label{example:hi-multimethod}
\end{exmp}

@@

Dispatching based on literal values enables certain forms of method
definition, but this is only part of the story for multimethod dispatch.

\paragraph{Class-based dispatch}
For class values, multimethods can choose methods based on subclassing
relationships.

Recall the multimethod from figure~\ref{fig:ex1}, reproduced here.
\begin{minted}{clj}
(ann pname [(U File String) -> (U nil String)])
(defmulti pname class)
(defmethod pname String [s] (pname (new File s)))
(defmethod pname File [f] (.getName f))
\end{minted}

%Its dispatching function is
%\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
%respectively.
\noindent
The dispatch function \clj{class}
%---associated at multimethod creation with \clj{defmulti}---
dictates 
whether the \clj{String} or \clj{File} method is chosen.
%---both installed via \clj{defmethod}
%
The multimethod dispatch rules use
\clj{isa?}, a hybrid predicate which is a subclassing check for classes and
an equality check for other values.

@@

In our example,
  \clj{(pname "STAINS/JELLY")}
chooses the  method for \clj{String} because
\clj{(isa? (class "STAINS/JELLY") String)}
is true and
\clj{(isa? (class "STAINS/JELLY") File)}
is false. The \clj{String} method body
\clj{(pname (new File "STAINS/JELLY"))}
chooses the \clj{File} method for opposite reasons,
resulting in 
\begin{minted}{clojure}
(.getName (new File "STAINS/JELLY")) ; :- (U nil Str)
;=> "JELLY"
\end{minted}

@@

Beyond primitives and Java objects, the most common Clojure data
structure is the 
immutable hash-map, typicially with keyword keys. This structure is
the primary way to represent compound data in Clojure programs.

Hash-maps are accessed with the \clj{get} function:
\begin{minted}{clojure}
(def breakfast {:en "waffles" :fr "croissants"})

(get breakfast :en)    ; :- Str
;=> "waffles"
\end{minted}
Additionally, keywords are functions that look themselves up in a
map. Missing keys produce \clj{nil}.
\begin{minted}{clojure}
(:fr breakfast)        ; :- Str
;=> "croissants"
(:bocce breakfast) ; :- nil
;=> nil
\end{minted}

In Typed Clojure, \emph{HMap types} describe the most common usages of
keyword-keyed maps.
\begin{minted}{clojure}
breakfast ; :- (HMap :mandatory {:en Str :fr Str}
          ;          :complete? true)
\end{minted}
The inferred type for \clj{breakfast} holds two kinds of information---the known entries--\clj{:en}
and \clj{:fr}---and their types, which are  \clj{:mandatory}, and that 
no other key is present, since \clj{:complete?} is \clj{true}.

HMap types default to  partial specification---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
\clj{'{:en Str :fr Str}}
omits
information about absent keys, only providing information on
\clj{:mandatory} keys.

@@

\paragraph{HMaps in practice} The next example is extracted from a production system at CircleCI,
a company with a large production Typed Clojure system
(section~\ref{sec:casestudy} presents a case study and empirical
result from this code base).

\begin{exmp}
\begin{minted}{clojure}
(defalias RawKeyPair
  "Unencrypted keypair -- extra keys disallowed"
  (HMap :mandatory {:public-key RawKey,
                    :private-key RawKey},
        :complete? true))

@@

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair "Encrypt an unencrypted keypair"
  [{pk :private-key :as kp}] ; original map is kp
  (assoc 
    ; remove unencrypted private key
    (dissoc kp :private-key)
    ; add encrypted private key
    :enc-private-key (encrypt pk)))
\end{minted}
\label{example:circleci}
\end{exmp}

%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
Using ully specified HMaps
statically enforce the private key is not accidentally left in a supposedly
encrypted keypair.

\begin{minted}{clojure}
(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
(defn enc-keypair-bad
  [{pk :private-key :as kp}]
  (assoc kp :enc-private-key (encrypt pk)))
  ; Type Error:
  ;   Expected EncKeyPair, given 
  ;   (HMap :mandatory {:enc-private-key EncKey
  ;                     :private-key RawKey
  ;                     :public-key RawKey}
  ;         :complete? true)
\end{minted}
The extra \clj{:private-key} entry does not match \clj{EncKeyPair}, so a type error
is raised.

@@

Since HMaps are the primary way of specifying the structure of data in Clojure,
and multimethods are the primary tool for dispatching on data, they are inevitably
linked.
There are infinite ways of both structuring and dispatching on data,
so we cannot hope to merely add a set of special case rules for
handling these features. Instead, as type system designers, we must
search for a compositional approach.

Thankfully, occurrence typing, originally designed for reasoning about
\clj{if} tests, provide the compositional approach we need.
By extending the system with
a handful of rules based on HMaps and other functions, 
we can automatically cover both easy cases and those
that compose simple rules in arbitrary ways.

Futhermore, this approach extends to multimethod dispatch---the primitive branching
mechanism works like the humble \clj{if} conditional. 
Only a small number of rules are needed
to encode the \clj{isa?}-based dispatch, themselves made of simple pieces.
In practice, this means that conditional-based control flow typing
 extends to  multimethod dispatch, and vice-versa.

@@

The \clj{:combo} method is verified to only structurally recur
on \clj{Order}s. This is achieved because we learn the argument---\clj{o}---must % em dash needs space here, looks weird
be of type
\clj{'{:Meal :combo}}
since
\clj{(isa? (:Meal o) :combo)}
must be true. Combining \clj{'{:Meal :combo}}
with the fact that \clj{o} is an \clj{Order}
eliminates possibility of \clj{:lunch} and \clj{:dinner}
orders, simplifying \clj{o} to
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
which is an appropriate argument for the recursive call.

@@

In the \clj{Long} method, Typed Clojure learns that
its argument is at least of type \clj{'{:desserts Long}}---since
\begin{minted}{clojure}
(isa? (class (:desserts o)) Long)
\end{minted}
must be true.
%
Knowing \clj{o} is also an
\clj{Order}, we can eliminate the possibility of
a \clj{:combo} meal, and deduce the type:
\begin{minted}{clojure}
(U '{:Meal ':lunch :desserts Int}
   '{:Meal ':dinner :desserts Int})
\end{minted}
Looking up \clj{:desserts} on this type returns an \clj{Int}.

In the \clj{nil} case, we know
\begin{minted}{clojure}
(isa? (class (:desserts o)) nil)
\end{minted}
is true---which implies \clj{(class (:desserts o))} is \clj{nil}.

Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
  \item \clj{o} is missing a \clj{:desserts} entry.
\end{itemize}
Equivalently, we learn \clj{o} is at least of type
\begin{minted}{clojure}
(U '{:desserts nil}
   ; :absent-keys, a set of known absent entries
   (HMap :absent-keys #{:desserts}))
\end{minted}
Since \clj{o} is originally an \clj{Order}, this simplifies to a \clj{:combo} meal, 
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
thus allowing both recursive calls to type check.

\paragraph{From multimethod to multiple dispatch}
Clojure multimethod dispatch, and Typed Clojure's handling of it, goes
even further, supporting dispatch on multiple arguments via vectors.
%
Dispatch on multiple arguments is beyond the scope of this paper,
but the same intuition applies---adding support for multiple dispatch
automatically allows arbitrary combinations and nestings
of it and previous simple dispatch rules.
